Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-27 2-Abstract.tex
sync

Model federation is a multi model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Our solution is fully implemented with the Openflexo framework and its internal Flexo Modeling Language (FML). It fulfills all the challenge requirements.

\noteSM{J'ai fait un premier essai d'abstract... un peu trop vendeur peut-etre?}
\noteJC{bof, pas de manière éhontée : il faut qu'un abstract soit un peu
  vendeur et les exigences sont bien remplies. La fin est très positive, mais
afactuelle selon moi.}
\noteSylvain{Ce qui m'embête un peu avec cet abstract c'est qu'on met en avant la fédération de modèles alors que sur ce cas précis on en fait pas vraiment (sauf pour la partie outillage avec les éditeurs diagrammatiques). Concernant la fédération de modèle proprement dite, je crois qu'il faut qu'on montre qu'on a besoin d'une couche conceptuelle (qu'on relie à l'espace technique via les connecteurs techniques, les model slots et les rôles). C'est ce besoin qu'adresse le language FML. On montre ici plutôt que le language FML - utilisé comme language conceptuel - est assez intéressant, assez expressif pour capturer les exigences du challenge d'une part, et va jusqu'à permettre l'exécution des processus. Et cerise sur le gâteau: on peut construire un outillage graphique pour editer et exécuter ce use case.}
\noteSylvain{En revanche, on peut arguer que pouvoir au même niveau manipuler modèle et métamodèle nous simplifie grandement les choses pour adresser des problématiques multi-niveaux > un bout de la conclusion}.
\noteSM{En fait l'idee de l'abstract c'est juste de nomer les approche utilisé. Les détails sont decrites dans la section 2}

>>>>

Model federation is a multi model \emph{management} approach based on the use of virtual models and loosely coupled links. The models in a federation remain autonomous and represented in their original technological spaces whereas virtual models and links serve as control components used to present different views to the users and  maintain synchronization. In this paper we tackle the \emph{process modeling challenge}, which consists in providing a solution to the problem of specifying and enacting processes. Solutions must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We present a solution based on model federation and discuss the advantages and limitations of using this approach for multi-level modeling. Concretely, we use virtual models and more precisely the Flexo Modeling Language (FML) that serves to build them as the main building block for the process modelling challenge whereas the federation aspect is used in order to provide editing tooling for the process language. Our solution is fully implemented with the Openflexo. It fulfills all the challenge requirements.

\noteSM{J'ai fait un premier essai d'abstract... un peu trop vendeur peut-etre?}
\noteJC{bof, pas de manière éhontée : il faut qu'un abstract soit un peu
  vendeur et les exigences sont bien remplies. La fin est très positive, mais
factuelle selon moi.}
\noteSylvain{Ce qui m'embête un peu avec cet abstract c'est qu'on met en avant la fédération de modèles alors que sur ce cas précis on en fait pas vraiment (sauf pour la partie outillage avec les éditeurs diagrammatiques). Concernant la fédération de modèle proprement dite, je crois qu'il faut qu'on montre qu'on a besoin d'une couche conceptuelle (qu'on relie à l'espace technique via les connecteurs techniques, les model slots et les rôles). C'est ce besoin qu'adresse le language FML. On montre ici plutôt que le language FML - utilisé comme language conceptuel - est assez intéressant, assez expressif pour capturer les exigences du challenge d'une part, et va jusqu'à permettre l'exécution des processus. Et cerise sur le gâteau: on peut construire un outillage graphique pour editer et exécuter ce use case.}
\noteSylvain{En revanche, on peut arguer que pouvoir au même niveau manipuler modèle et métamodèle nous simplifie grandement les choses pour adresser des problématiques multi-niveaux > un bout de la conclusion}.
\noteSM{En fait l'idee de l'abstract c'est juste de nomer les approche utilisé. Les détails sont decrites dans la section 2-Tecnology. On peut nommer FML avant mais je crois que ça serait moins claire. Cela dit, n'hesitez pas a modifier.}

%The models in a federation remain autonomous and represented in their original technological spaces %whereas virtual models and links serve as control components used to present different views to the %users and  maintain synchronization.

### 2021-05-27 6-ModelPresentation.tex
sync

Figure \ref{fig:ToolingArchitecture} shows the architecture of these two tools. Left of figure presents process type graphical editor. ProcessTypeEditor is implemented as a VirtualModel declaring two model slots (represented with bold circles). The first model slot references an instance of Acme metamodel, while the second model slot references an instance of diagram, conforms to the ProcessType diagram specification (a diagram metamodel

@@

>>>>

@@

Figure \ref{fig:ToolingArchitecture} shows the architecture of these two tools. Left of figure presents \textit{process type graphical editor}. \textit{ProcessTypeEditor} is implemented as a \textit{VirtualModel} declaring two model slots (represented with bold circles). The first model slot references an instance of Acme metamodel, while the second model slot references an instance of diagram, conforms to the \textit{ProcessType diagram specification} (a diagram "metamodel" which defines and specify structure and graphical representations for edited items). While executed, this tool manages a graphical view for an instance of Acme metamodel (the \textit{Acme model}) and a specific \textit{ProcessType} instance. This tools allows to represent and edit a \textit{ProcessType}. One particular point should be noted regarding the highly reflective nature of FML and associated tooling : when drag and drop interactor is applied for a new item, the tool allows to choose the concept type to be instantiated (when a \textit{TaskType} is created for example, the user must choose the sub-concept of \textit{TaskType} to be selected - in Acme use case it can be \textit{SETaskTask} or \textit{CodingTaskType} or default value \textit{TaskType}).

The other tool, an \textit{enacted process editor}, provides process edition and tasks assignations through a graphical visualization displaying process being executed. This tool is represented one the right side of the figure \ref{fig:ToolingArchitecture}. It follows the same architecture pattern as explicited for the \textit{process type graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \textit{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \textit{TaskType} definition, and required assignments applies (if for example "Coding" \textit{TaskType} defines COBOL as output programming language, related task is automatically assigned to "Ann Smith"). All tasks get a status (\textit{Not startable} - when not assigned or when required input artifacts are not available - , \textit{Startable}, \textit{Started}, \textit{Completable}, \textit{Completed}), as well as a set of performing actors, a begin and end date, and used and produced artifacts. This tools also manages creation of artifacts, whose semantics follows rules defined in section \ref{sec:ProcessEnactment}}
